This is the title [Article v\versionnumber]

### 2022-10-05 manuscript/manuscript.tex
Start of the restructuring process
Changed some wording to reflect the split into a bsaic and advanced
tutorial
Added some tables to the Running GIST section for more information
concerning cpptrajs gist option.

\usepackage{listings}

@@

This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.

Furthermore, a GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
Using this test system, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and work with their output data.
Previous experience of GIST is not necessary.
However, we only provide a short explanation of the theory behind GIST and point the reader towards the primary sources for more detailed information.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} as part of AmberTools \cite{amber22}.

A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).
The closed-source Amber simulation engine is used in this tutorial to run the necessary MD simulations.

@@

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.
\section{Theory}

GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al. \cite{Ramsey2016}. % to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
Here, we only provide a short overview of the theory behind GIST.
For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.

\section{Running GIST}
%\subsection{Running the GIST analysis}
To run a gist analyis, the following is required:
\begin{itemize}
	\item A running installation of \software{cpptraj}.
	\item A topology of the molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

@@

\paragraph{Postprocessing and Analyzing GIST results}
Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a full list of quantities calculated by GIST \todo{This is incomplete!}:

\begin{itemize}
	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. For water, gO and gH are produced.}
	%\item{[gH] Number density of hydrogen centers found in the voxel in units of the reference bulk density.}
	\item{[Esw] Mean solute-water interaction energy density.}
	\item{[Eww] Mean water-water interaction energy density.}
	\item{[PME] (only if PME was used) water PME energy density.}
	\item{[dTStrans] First order translational entropy density.}
	\item{[dTSorient] First order orientational entropy density .}
	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
	\item{[dipole] Magnitude of mean dipole moment (polarization).}
	\item{[order] (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter.}
	\item{[dipolex] x-component of the mean water dipole moment density}
	\item{[dipoley] y-component of the mean water dipole moment density}
	\item{[dipolez] z-component of the mean water dipole moment density}
	\item{[Eij] (if \inlinecode{doeij} was specified) Water-water interaction matrix.}
\end{itemize}

Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen. All quantities mentioned above are written to the output file (output.dat) and can be used to generate DX files for other quantities.

\section{Tutorial}
Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
In the process, the reader will learn how to set up and run a GIST calculation.
During the tutorial, we will
\begin{enumerate}
	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
\end{enumerate}

@@

\subsection{Tutorial data}
We will use the 1STP crystal structure to start our calculations.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
\subsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

\subsection{Running MD for GIST}

@@

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
\end{figure}

>>>>

\usepackage{listings}
\usepackage{tabularx}

@@

This tutorial provides an example of a GIST workflow that can be easily adapted towards different systems. 
In addition to a short introduction of the method and the theory behind it, we present an overview of the various different flavors of GIST methods.
We ease the user into the basic usage of GIST by presenting a calculation for a small molecule.
Building on the basics, a more advanced GIST study of the streptavidin-biotin complex is presented, which reflects on many common aspects of GIST studies.
Using these test systems, we present solvation analysis of a small molecule (biotin) as well as a biomolecule (streptavidin).
We show how to interpret the three-dimensional contributions of solvation free energy in a binding pocket and around the ligand.

@@

\subsection{Background knowledge}
This tutorial is aimed at users with a solid knowledge of molecular dynamics (MD) simulations.
The user should be able to run MD simulations and work with their output data.
Previous experience of GIST is not necessary.
However, we only provide a short explanation of the theory behind GIST in section \ref{sec:theory} at the end of this tutorial. For more detailed information, we point the reader towards the primary sources for more detailed information.

@@

\subsection{Software/system requirements}
%Tutorials should clearly define what system and/or software requirements the researcher will need to complete the tutorial (e.g., VMD version 1.9 or newer, AMBER, etc.). Tutorials requiring specific software packages must provide instructions and files for the referenced version of the software.
The main implementation of GIST is part of the MD analysis software \software{cpptraj} released with AmberTools \cite{amber22}.

A recent version of \software{cpptraj} should be used, such as the one provided in AmberTools22 or a recent installation from GitHub (\url{https://github.com/Amber-MD/cpptraj}).

@@

For the visualization of three-dimensional contributions we require molecular visualization software that can handle \software{OpenDX} files.
We recommend using a current version of \software{PyMol}\cite{pymol} or \software{VMD}\cite{vmd}.

\section{Running GIST}
%\subsection{Running the GIST analysis}
To run a GIST analyis, the following is required:
\begin{itemize}
	\item A running installation of \software{cpptraj}.
	\item A topology of the molecule of interest in a solvent box.
	\item An MD trajectory of that system.
\end{itemize}
GIST analysis can be divided in three steps: defining a region of interest, running GIST using \software{cpptraj}, and post-processing and analysis.
\cite{Ramsey2016}

@@

Cpptrajs' gist action has a large number of options to modify the GIST calculation or its' output. A full list is provided at the end of this tutorial or through cpptrajs manual. The major options are summarized in the following table:

\begin{table}[h]
	\caption{The major options available in cpptrajs' GIST implementation.}\label{tab:GIST_options}
	\small
	\begin{tabularx}{\columnwidth}{@{}llXl@{}}
		\toprule
		Command       & Options & Explanation & Default  \\
		\midrule
		gridcntr    & <y> <y> <z> & Coordinates of the grid center in \AA & 0 0 0  \\
		griddim     & <nx> <ny> <nz> & Number of voxels along each axis defining the grid size & 40 40 40\\
		gridspacn & <dval> &Grid spacing / voxel dimensions in  \AA & 0.5    \\
		refdens         & <rdval> & Reference density of the bulk water in molecules/\AA$^3$ . & 0.0334    \\
		\bottomrule
	\end{tabularx}
\end{table}

Setting the reference density according to the water model used in your simulation is important to get correct results. The reference density can be calculated from a GIST calculation for a pure water simulation or estimated from bulk-like water sufficiently far away from a solvent. \todo{How much information on the choice of reference density should we include?}

\paragraph{Postprocessing and Analyzing GIST results}
Cpptraj outputs the quantities calculated as a "open data explorer" (.dx) file format which can be easily visualized in Pymol or VMD.
Aside from the thermodynamic quantities, GIST also calculates the number densities of hydrogen and oxygen in each voxel. Here is a list of the most important quantities calculated by GIST:

\begin{itemize}
	\item{[gX] For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. For water, gO and gH are produced.}
	\item{[Esw] Mean solute-water interaction energy density.}
	\item{[Eww] Mean water-water interaction energy density.}
	\item{[PME] (only if PME was used) water PME energy density.}
	\item{[dTStrans] First order translational entropy density.}
	\item{[dTSorient] First order orientational entropy density .}
	\item{[dTSsix] First order entropy density for the six-dimensional combined orientation and translation space.}
	\item{[neighbor] Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density.}
	\item{[dipole] Magnitude of mean dipole moment (polarization).}
	\item{[dipolex] x-component of the mean water dipole moment density}
	\item{[dipoley] y-component of the mean water dipole moment density}
	\item{[dipolez] z-component of the mean water dipole moment density}
\end{itemize}

Several dx files are written by default, such as the energies, entropies, number density of oxygen, and number density of hydrogen. All calculated quantities are written to the output file (output.dat) and can be used to generate DX files for other quantities. A full list of all calculated properties found in the output in addition to the ones mentioned above is provided at the end of the tutorial.

\section{Tutorial}
Here, we aim to guide the reader through a GIST analysis of the streptavidin-biotin complex.
During the tutorial, we will
\begin{enumerate}
	\item set up and run GIST calculations for a small molecule as well as a protein binding pocket.
	\item compute the free energy of solvation for a small molecule using GIST.
	\item visualize local contributions to $\dgsolv$ in the streptavidin binding pocket as well as around biotin.
	\item compute the binding free energy of the complex from the individual energy contributions combined with the contribution of hydration computed using GIST.
	\item explore advantages and disadvantages of the GIST method when used to calculate  $\dgsolv$
\end{enumerate}
The tutorial is divided into a beginners part and an advanced part. In the beginners part, we will take a look at how to run GIST calculations for small molecules and interpret the output for a small molecule. In the advanced part, we will apply this knowledge to a protein binding pocket, visualize $\dgsolv$ hotspots and estimate a free energy of binding for a small molecule into this pocket using an end-state approach. In addition, while the beginners part focuses only on GIST calculation run, we will discuss all steps along the way in detail in the advanced section.

@@

\subsection{Tutorial data}
We will use the 1STP crystal structure to start our calculations. For the beginners section, we will use the biotin ligand as extracted from the crystal structure. For the second part, we will use the whole crystal structure.
For the sake of simplicity, you can download prepared and solvated Amber topologies and structures based on 1STP by running
\begin{lstlisting}[style=bash]
git clone git@github.com:liedllab/gist-tutorial.git
\end{lstlisting}
This will download the tutorial files (including both the manuscript and the examples in the \software{code} folder).
\
\subsection{Running GIST for Biotin}

\subsection{Applying GIST for the Streptavidin-Biotin complex}
\subsubsection{System Preparation and Equilibration}
If you want to skip this section, run 
\begin{lstlisting}[style=bash]
make equilibration-targets
\end{lstlisting}
in the code folder. This assumes that you have Python and Amber set up properly and that \software{pmemd.cuda} is in the PATH.

@@

\subsubsection{Running MD for GIST}

@@

%\begin{figure}
%	\centering
%	\includegraphics[width=1.0\linewidth]{streptavidin_bp.png}
%	\caption{Volume of strepatividin defined around 3.5 A of biotin. }\label{streptavidin_sasa}
%\end{figure}

\section{Theory}
\label{sec:theory}
GIST is an implementation of inhomogeneous fluid solvation theory (IST) \cite{Lazaridis1998} that discretizes the free energy of solvation $\dgsolv$ on a three-dimensional grid. 
It was first devised by Nguyen et al. \cite{Nguyen2012} and its implementation in \software{cpptraj} was thoroughly described by Ramsey et al. \cite{Ramsey2016}. % to overcome the limitation of IST which was commonly used for regions of high water occupancy and its implementation in \software{cpptraj} was thoroughly described in \cite{Ramsey2016}.
Here, we only provide a short overview of the theory behind GIST.
For more detailed information, we recommend one of the more recent publications on developments in GIST \cite{Kraml2020,Chen2021}.
